\documentclass[11pt,a4paper]{report}

\usepackage[english]{babel}

\usepackage[hidelinks,unicode]{hyperref}
\hypersetup{
    pdftitle={Advanced Operating Systems - Report},
    pdfauthor={
        Fabian Wüthrich,
        Janis Peyer,
        Kristina Martšenko,
        Philippe Mazenauer
    }
}

\usepackage{csquotes}
\usepackage{biblatex}
\addbibresource{main.bib}

\usepackage{listings}
\usepackage{bytefield}
\usepackage{xcolor}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},   
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,         
    breaklines=true,                 
    captionpos=b,                    
    keepspaces=true,                 
    numbers=none,                    
    numbersep=5pt,                  
    showspaces=false,                
    showstringspaces=false,
    showtabs=false,                  
    tabsize=2
}
\lstset{style=mystyle}

% For sequence diagram
\usepackage{tikz}
\usepackage{pgf-umlsd}
\usepgflibrary{arrows}

\begin{document}

\title{Advanced Operating Systems - Report}
\author{
Fabian W{\" u}thrich
\and
Janis Peyer 
\and
Kristina Mart{\v s}enko
\and
Philippe Mazenauer
}
\date{Spring 2020}

\maketitle

\tableofcontents

\listoffigures

\listoftables

\clearpage

% One chapter, but easier to handle this way
\include{memory}

\include{paging}

\input{milestone2.tex}

\input{milestone3.tex}

\input{milestone4}

\input{milestone5}

\input{ump}

\input{nameserver}

\input{network}

\input{filesystem}

\input{shell}

\chapter{Summary of overall system}

It proved to have been good decision to invest time in writing test code
(see \verb|lib/grading|, but also other places like \verb|usr/memeater/main.c|,
\verb|usr/test/filereader/main.c|).
Because during the intergration we did not encounter further problems with the
core system functionality like paging and memory allocation.

We did not end up with as a modular system as imagined in the beginning, with
everything being its own service in its own domain.
It's quite tempting and also prudent given the little time that if something
works like the memory server, to not refactor it into it's own domain.

So init still does all of the core os work, and only the network project managed
to be truly in it's separate domain.

Especially in the rpc part, it felt sometimes more like programming for a
distributed system than just simply for a multicore processor.
All the many things that happen on Linux hiddenly are much more in present, and
that's a good thing.
One much more concerns one self with the reality and limitations of the system
this way.
Which is good, because it allows one solve the problem in a way that's
appropriate to the problem, not having to rely on that the abstractions have
enough holes in them to not stand in the way.

Even though it operates on radically different principles, because the
principles are so elegant, simple and more importantly, appropriate to todays
problems, one feels very much at home even after this short a time.

\appendix

\input{userguide}

\printbibliography

\end{document}
